Analysing Constraint Grammars with a SAT-solver

### 2016-03-02 paper/lrec/eval_futurework.tex
stuff

We tested three grammars to find conflicting rules: 
Dutch\footnote{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-nld/apertium-nld.nld.rlx}},
with 59 rules; 
Spanish\footnote{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}},
with 279 rules; and 
Finnish\footnote{\url{https://github.com/flammie/omorfi/blob/master/src/vislcg3/omorfi.cg3}},
with 1185 rules. We left out \textsc{add}, \textsc{map} and other rule
types introduced in CG-3, and only tested \textsc{remove} and \textsc{select} rules.
The results for Dutch and Spanish are shown in table~\ref{table:res},
and the results for Finnish in table~\ref{table:resFin}.

A natural follow-up evaluation would be to compare the performance of the
grammar in the original state, and after removing the conflicts found
by our tool. Unfortunately, we did not have time to perform such
evaluation, and in addition, we only have gold standard corpus (20~000) words for Spanish.

@@

\hline
                   & \textsc{nld}  & \textsc{spa}lem  & \textsc{spa}nolem \\ \hline
\# rules           & 59            & 279       & 279     \\ \hline
\# readings        & 336           & 3905      & 1735    \\ \hline
\# conflicts       & 7             & n         & m    \\ \hline
\clock{} all rules & ??s           & + 1h      & 30m-1h   \\ \hline
% \clock{} last rule & ??s           & ??s       & ?min ??s    \\ \hline
\clock{} no amb. 
classes            & 10ish sec       & ??min ??s    & ?h ??min    \\ \hline

@@

\end{tabular}
\caption{\todo{make more tables}}
\label{table:res}
\end{table}

\hline
              & n clitics & m clitics & only underspecified \\ \hline
\# rules      & 1185      & 1185      & 1185\\ \hline
\# readings   & many & more & only like <2000   \\ \hline
\# conflicts  & n   & m        & super many    \\ \hline
\clock{} all rules       & ??s              & ??min ??s    & ?h ??min    \\ \hline
\clock{} last rule       & ??s              & ??s          & ?min ??s    \\ \hline

\end{tabular}
\caption{Special table for Finnish}
\label{table:resFin}
\end{table}

The experiments revealed problems in all grammars. For the smaller
grammars, we were able to verify manually that the detected rules were
true positives.
%The results for the Finnish grammar are inconclusive, 
% For the Finnish grammar, we did not have time to
% investigate all of them, but among those we could, we found both true and false negatives.
We did not systematically check for false negatives in any of the
grammars, but we kept track of a number of known tricky cases; mostly
rules that had been false positives in earlier stages of the
development, and other rules with similar patterns, as well as rules
with negations.
%While this does not guarantee the lack of false negatives, it indicates some degree of reliability. 
As the tool matures and we add
new features, a more in-depth analysis will be needed.

\paragraph{Dutch} The Dutch grammar had two kinds of errors: rule-internal, and rule interaction. As for rule-internal conflicts, one was due to a misspelling in the list definition for personal pronouns, which rendered 5 rules ineffective. The other was about subreadings: the genitive \emph{s} is analysed as a subreading in the Apertium morphological analyser, but it appeared in 2 rules as the main reading.

@@

@@

\paragraph{Spanish} The Spanish grammar had proportionately the highest number of errors. \todo{confirm; I assume this is gonna be so, but let's see Fin and Por}.  The grammar we ran is like the one found in the Apertium repository, apart from 2 changes: we fixed some typos (capital O for 0) in order to make it compile, and commented out two rules that used regular expressions, because our solution does not handle them properly (see more on Future work/Implementation).

@@

In addition, there was a number of definitions that were never
used. The program doesn't currently point them out to the user, but
that is a trivial addition; just like rule-internal conflicts, it does
not require any semantic analysis. 
We also discovered problems in some set definitions, when we tried to
use the set definitions from the grammar directly as our readings.
For instance, the following definition requires the word to be all of
the listed parts of speech at the same time:
\begin{itemize}
\item[] 
\texttt{SET NP\_Member = N + A + Det + PreAdv + Adv + Pron ;}
\end{itemize}

This particular definition was never used in the rules, so it would not
have caused a conflict. We noticed it by accident, when the SAT-solver
offered it in an example sequence that, meant for another rule.

@@

We found the same kind of formulation in many other rules and grammars.
To catch them more systematically, we could add a feature that alerts in all cases where a condition with 0 is used, and automatically construct a version of the rule that moves whatever tags in the 0-condition into the target, then asks the user which one was meant.

@@

The comments make it clear that the first rule is meant to disambiguate between poder and podar, and the second rule between sentir and sentar. But the rule does not mention anything about ``poder'' nor ``sentir''; the SAT-solver gave a triggering sequence without

This example shows a shortcoming of our way of handling the lemmas. If
the lemmas and word forms were integrated to the analyses in a more
feasible way, then the ambiguity class constraints would create a
sequence where ``podar'' was ambiguous with ``poder'', instead of randomly chosen readings. So, even though this was not marked as a conflict, we can say that the weird output of the tool was unnecessary, and by better implementation, the tool would have created more realistic example sequence.

\paragraph{Finnish}

\todo{Rerun the tests and rewrite this whole section!}

@@

In addition to the previous problem, we had a large number of readings. In order to make the test run faster, we tried a few shortcuts. First, we restricted the amount of clitics per word form, and even ignored all the specified readings

The Finnish grammar was the largest and most complex of all the
grammars we have tested.

To improve performance, we ignored the fully specified readings from the
lexicon and used only tag combinations from the grammar:
for example, the rule \texttt{REMOVE (verb sg) IF (-1 det)}
gives us ``verb sg'' and ``det''.
The shortcut works most of the time, but it is possible to
miss some cases: e.g. \texttt{SELECT Pron + Rel IF (0 Nom)} 
may assume ``pron rel nom'' in one reading, but our method only gives
``pron rel'' and ``nom'' separately. 
17 false positives were caused by this shortcut. 
In addition, 2 false positives were caused by the naive implementation of
rules with \verb!*! in their context.

Due to these reasons, the results about the Finnish grammar are
inconclusive. We only include them to show some preliminary data on larger grammars.
%For a condition situated $n$ or more words to the left or right, we only create a word exactly $n$ words away. This is something to be fixed in the future.

\paragraph{Performance} The speed of the check is an obvious issue,
especially for the Finnish grammar. 
As mentioned before, we used 2000 readings extracted from the grammar.
With the full set of 6000 readings from the lexicon, the speed
was even worse: checking the last rule took 30 minutes, as opposed to
3 minutes with 2000 readings. 
%We never finished running the whole rule set with 6000 readings, but
%it would have taken potentially days.
We believe we can still signifcantly improve the performance of the SAT-encoding. Currently it is just a naive proof-of-concept implementation, and we have not tried to optimise for speed at all.

@@

The evaluation indicates that the tool finds cases that are not
obvious to the human eye.

One additional feature is to suggest reformattings for a rule. Recall
figure~\ref{fig:regroup} from the introduction; in that case, the
original rule was written by the original author, and another
grammarian thought that the latter form is nicer to read. Doing the
reverse operation could also be possible. If a rule with long
disjunctions conflicts, it may be useful to split it into smaller
conditions, and eliminate one at a time, in order to find the
reason(s) for the conflict.

Our solution to hardcode the tag combinations in the readings is feasible for simple morphology, but it can cause problems with more complex morphology. One big downside is that in order to implement \textsc{ADD}, \textsc{ADDREADING} and \textsc{MAP}, we need to be prepared for new readings--even if the lexicon gives all readings that exist in the lexicon, the user might give a nonexistent reading, or in the case of MAP, a syntactic tag, which is (by definition) not in the lexicon. A more scalable solution would be to make each tag a variable, and ask the question ``can this reading be a noun? how about singular? how about conditional?'' separately for each tag. Then we could lift the restriction of tag combinations into a SAT solver: make SAT clauses that prohibit a comparative to go with a verb, or conditional with a noun.
Another benefit of this solution is the addition of lemmas and word forms, as well as regular expressions: currently, if we want to add one more lemma for a verb, we need to create as many new variables as there are distinct verb forms---easily hundreds for languages with rich morphology.
\todo{less rambling: For rules with regular expressions, this would blow up even more: there are rules that only address whether the word form in a condition starts with a lowercase letter, or whether it ends in a certain suffix. A realistic, albeit unsatisfactory option is just to treat it as an underspecified reading, and offer that same regular expression to the user when generating a symbolic sentence. Another, less feasible, solution is to grep all the word forms or lemmas that match the regular expression in question---in the worst case (e.g. lower case), this will match all the entries in the expanded lexicon.}

@@

>>>>

We tested three grammars to find conflicting rules: 
Dutch\footnote{\scriptsize{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-nld/apertium-nld.nld.rlx}}},
with 59 rules; 
Spanish\footnote{\scriptsize{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}}},
with 279 rules; and 
Finnish\footnote{\scriptsize{\url{https://github.com/flammie/apertium-fin/raw/master/apertium-fin.fin.rlx}}},
with 1185 rules. We left out \textsc{add}, \textsc{map} and other rule
types introduced in CG-3, and only tested \textsc{remove} and \textsc{select} rules.
The results for Dutch and Spanish are shown in table~\ref{table:res},
and the results for Finnish in table~\ref{table:resFin}.

A natural follow-up evaluation would be to compare the performance of the
grammar in the original state, and after removing the conflicts found
by our tool. Unfortunately, we did not have time to perform such
evaluation, and in addition, we only have gold standard corpus (20~000~words) for Spanish.

@@

\hline
                   & \textsc{nld}  & \textsc{spa}lem  & \textsc{spa}nolem \\ \hline
\# rules           & 59            & 279       & 279     \\ \hline
\# readings        & 336           & 3905      & 1735    \\ \hline
\# true positives  & 7             & n         & m    \\ \hline
\# false positives & 0             & n         & m    \\ \hline
\clock{} with amb. 
           classes & ??s           & + 1h      & 30m-1h   \\ \hline
% \clock{} last rule & ??s           & ??s       & ?min ??s    \\ \hline
\clock{} no amb. 
           classes & 10ish sec       & ??min ??s    & ?h ??min    \\ \hline

@@

\end{tabular}
\caption{Results for Dutch and Spanish grammars.}
\label{table:res}
\end{table}

The experiments revealed problems in all grammars. For the smaller
grammars, we were able to verify manually that the detected rules were
true positives.
%The results for the Finnish grammar are inconclusive, 
% For the Finnish grammar, we did not have time to
% investigate all of them, but among those we could, we found both true and false negatives.
We did not systematically check for false negatives in any of the
grammars, but we kept track of a number of known tricky cases; mostly
rules with negations and complex set operations.
%While this does not guarantee the lack of false negatives, it indicates some degree of reliability. 
As the tool matures and we add new features, a more in-depth analysis
will be needed.

\subsection{Dutch} The Dutch grammar had two kinds of errors: rule-internal, and rule interaction. As for rule-internal conflicts, one was due to a misspelling in the list definition for personal pronouns, which rendered 5 rules ineffective. The other was about subreadings: the genitive \emph{s} is analysed as a subreading in the Apertium morphological analyser, but it appeared in 2 rules as the main reading.

@@

We found the same kind of definitions in many other rules and grammars.
To catch them more systematically, we could add a feature that alerts in all cases where a condition with 0 is used, and automatically construct a version of the rule that moves whatever tags in the 0-condition into the target, then asks the user which one was meant.

@@

\subsection{Spanish} The Spanish grammar had proportionately the
highest number of errors. The grammar we ran is like the one
found in the Apertium repository, apart from two changes: we fixed
some typos (capital O for 0) in order to make it compile, and
commented out two rules that used regular expressions, because our
solution does not handle them properly; see more on Section~\ref{sec:conclusion}.

@@

In addition, there was a number of set definitions that were never
used. Since VISL CG-3 already points out unused sets, we did not add such
feature in our tool. However, we noticed an unexpected benefit when
we tried to use the set definitions from the grammar directly as our
readings: this way, we can discover inconsistencies even in
set definitions that are not used in any rule.
%If a set definition is impossible, using it in a rule makes the rule have an internal conflict.
For instance, the following definition requires the word to be all of
the listed parts of speech at the same time---the grammar writer meant
OR instead of +:
\begin{itemize}
\item[] 
\texttt{SET NP\_Member = N + A + Det + PreAdv + Adv + Pron ;}
\end{itemize}

If it was used in any rule, that rule would have been marked as
conflicting. We noticed the error by accident, when the program
offered the reading \texttt{w2<n><np><adj><det><preadv><adv><prn>}
in an example sequence meant for another rule.

@@

@@

The comments make it clear that the first rule is meant to disambiguate between poder and podar, and the second rule between sentir and sentar. But the rule does not mention anything about ``poder'' nor ``sentir''; the SAT-solver gave a triggering sequence without 
We found the same kind of formulation in many other rules and grammars.
To catch them more systematically, we could add a feature that alerts in all cases where a condition with 0 is used, and automatically construct a version of the rule that moves whatever tags in the 0-condition into the target, then asks the user which one was meant.

This example shows a shortcoming of our way of handling the lemmas. If
the lemmas and word forms were integrated to the analyses in a more
feasible way, then the ambiguity class constraints would create a
sequence where ``podar'' was ambiguous with ``poder'', only in the
word worm ``<puedo>'' instead of randomly chosen readings. So, even though this was not marked as a conflict, we can say that the weird output of the tool was unnecessary, and by better implementation, the tool would have created more realistic example sequence.

\begin{table*}[t!]
\centering
\begin{tabular}{|l|l|l|l|}

\hline
              & 1 clitic & 2 clitics & readings only from the grammar \\ \hline
\# readings   & many     & more & \textless{}2000   \\ \hline
\# true 
   positives  & n   & m        & o    \\ \hline
\# false 
   positives  & p   & q        & super many    \\ \hline
\clock{}amb.
      classes & ??s              & ??min ??s    & ?h ??min  \\ \hline
\clock{}no amb.
      classes & ??s              & ??s          & ?min ??s  \\ \hline

\end{tabular}
\caption{Results for Finnish (1185 rules).}
\label{table:resFin}
\end{table*}

\subsection{Finnish}

@@

In addition to the previous problem, we had a large number of
readings. In order to make the test run faster, we used a few shortcuts.
shortcuts. 
Firstly, we restricted the amount of clitics per word form, and even
ignored all the specified readings.
Secondly, we added ambiguity class constraints for only
\emph{intraparadigmatic ambiguities}: homonymous wordforms belonging
to the same lexeme, in a way that is systematic to the whole
paradigm. We reduced the morphological lexicon to contain only one
example of each inflection paradigm, and then expanded it: with all
the clitics, this resulted in \todo{n} word forms. Then we applied the
normal method of creating the ambiguity classes, and specified some
default cases manually, such as punctuation and word boundaries.
Thirdly, we ignored the fully specified readings from the
lexicon and used only tag combinations from the grammar, as described
in Section~\ref{sec:realistic_readings}.

The results of these three variants are described in
table~\ref{table:resFin}. As expected, using only readings from the
grammar resulted in most false positives.

\subsection{Performance} 
The running time of the grammars goes from seconds to hours. 
As can be seen from the times, increasing the number of readings
matters more than increasing the number of rules.
In addition, the ambiguity class constraints take a long time.

However, many of the use cases do not require running the whole
grammar. Testing the interaction between 5--10 rules takes just
seconds, regardless of the language.

%We believe we can still signifcantly improve the performance of the SAT-encoding. Currently it is just a naive proof-of-concept implementation, and we have not tried to optimise for speed at all.

@@

The evaluation indicates that the tool finds conflicts and dead rules
from actual grammars. We did not have a volunteer to test the tool in
the process of grammar writing, so we cannot conclude whether the
constructed examples are useful for getting new insights on the rules.
However, there are still a number of features to improve and add.

\paragraph{Combining morphological and lexical tags}

Our solution to hardcode the tag combinations in the readings is
feasible for simple morphology, but it can cause problems with more
complex morphology.
Another benefit of this solution is the addition of lemmas and word
forms, as well as regular expressions: currently, if we want to add
one more lemma for a verb, we need to create as many new variables as
there are distinct verb forms---easily hundreds for languages with
rich morphology.
This may also be feasible to support regular expressions, in a nicer
way than just spitting out the regex. (MORE ON THAT)

\paragraph{Full expressivity of CG-3}

In order to implement \textsc{substitute},
\textsc{add}, \textsc{addreading} and \textsc{map}, we need to be
prepared for new readings--even if the lexicon gives all readings that
exist in the lexicon, the user might give a nonexistent reading, or in
the case of MAP, a syntactic tag, which is (by definition) not in the
lexicon. A more scalable solution would be to make each tag a
variable, and ask the question ``can this reading be a noun? how about
singular? how about conditional?'' separately for each tag. Then we
could lift the restriction of tag combinations into a SAT solver: make
SAT clauses that prohibit a comparative to go with a verb, or
conditional with a noun.

% \paragraph{Regular expressions}
% For rules with regular expressions, this would blow up even more: there are rules that only address whether the word form in a condition starts with a lowercase letter, or whether it ends in a certain suffix. A realistic, albeit unsatisfactory option is just to treat it as an underspecified reading, and offer that same regular expression to the user when generating a symbolic sentence. Another, less feasible, solution is to grep all the word forms or lemmas that match the regular expression in question---in the worst case (e.g. lower case), this will match all the entries in the expanded lexicon.

\paragraph{Support for grammar writers}
One additional feature is to suggest reformattings for a rule. Recall
figure~\ref{fig:regroup} from the introduction; in that case, the
original rule was written by the original author, and another
grammarian thought that the latter form is nicer to read. Doing the
reverse operation could also be possible. If a rule with long
disjunctions conflicts, it may be useful to split it into smaller
conditions, and eliminate one at a time, in order to find the
reason(s) for the conflict.

@@

\subsection{Conclusions}

It is nice!
